---
title: "CDS 101 – Final Project Report"
author: "Group 4 / Ahmed D, Ahmed Hassan, John Stevenson"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "gmu-cds101.css"
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
---

> Replace this template text with your own writing, keeping the headings and overall structure aligned with the **CDS 101 project rubric**.

# 1. Problem Definition

- How many of the Tesla deaths were the result of a DUI from the year 2013 to 2022 in the United States? We separated the data to find out if DUIs make up more than 20% of the Tesla deaths. We are assuming that the dataset given is accurate and detailed.


# 2. Data Acquisition & Description

- Name of the data is Tesla Deaths (Updated 2023).
- Description of the accident weither they were DUI or other.
- Data was obtained from Kaggle link is: (https://www.kaggle.com/datasets/thedevastator/tesla-accident-fatalities-analysis-and-statistic).  
- Approximate size (296 rows, 24 columns).

```{r setup-data, echo=FALSE, message=FALSE, warning=FALSE}
# Example: load a dataset (replace with your own)
library(tidyverse)

library(dplyr)

library(readr)
Tesla_Deaths_Deaths_1_ <- read_csv("Tesla_Deaths_Deaths_1_.csv")

```

# 3. Data Cleaning & Preprocessing

Describe the steps you took to make the data usable, for example:

- We used string detection to convert the Description column into “DUI” and “Other” this made the grouping clear and easy to follow.
- We used a text classification model to model tesla deaths by DUI and non DUI. This allows for an easily reproducible method that is efficient. The baseline comparison allowed us to find if DUI deaths are common.

```{r cleaning}
New_Tesla <- Tesla_Deaths_Deaths_1_ %>%
mutate(
Dscrip = if_else(
str_detect(Description, "DUI"),
"DUI",
"other"
)
)
```

# 4. Exploratory Data Analysis (EDA)

This section maps directly to the **EDA** part of the rubric:

- Ggplot2 was used to perform visuals for our models.
- DUI cases 2.5% of Tesla deaths between between the years 2013-2022. 97.5% of all other accidents are due to other causes. There is an increase in both categories as time goes on due to more Teslas being driven.

```{r eda-summary, message=FALSE, warning=FALSE}
New_Tesla %>%
ggplot() +
geom_freqpoly(aes(x = Year, color = Dscrip, )) +
labs(
title = "DUI",
x = "year",
y = "Description"
)
```

# 5. Visualization Quality and Storytelling

Use this section to satisfy the **Visualization Quality** rubric criterion:

- We selected a frequency polygon (line graph) because the lines make it easy to see how the number of deaths changes over time and to compare DUI deaths to other deaths year by year.
- We chose this plot type because the lines are easier to read when looking for differences between the two groups across years.
- We used clear axis labels (Year on the x-axis and Number of Deaths on the y-axis) and a simple title so the viewer immediately understands what the plot is showing.
- The background is kept clean with minimal gridlines and readable text so the viewer can focus on the patterns in the data instead of extra decoration.

# 6. Modeling Approach

- We framed this as a binary classification problem: DUI vs. Other.
- We used a simple rule-based text model: check if “DUI” appears in Description.
- Our baseline idea was that DUI deaths might be about 20% of cases.
- This simple approach is transparent, easy to reproduce, and fits the small dataset.

# 7. Model Implementation & Evaluation

This corresponds to the **Model Implementation & Evaluation** rubric criterion.

Describe:

A text classification model will be the model we will be using for the project separating tesla death by DUI and non DUI deaths. Ggplot2 will be used to perform visuals for our models . We used the geom freqploy to  plot our graph for visuals.

```{r modeling, eval=FALSE}
New_Tesla %>%
  count(Dscrip) %>%
  mutate(percent = n / sum(n) * 100)
```

# 8. Conclusions & Recommendations

Summarize the key takeaways:

For our project we made a goal to find out how of the of the Tesla deaths were due to DUI. Our original hypothesis was the DUIs would make up 20% of the Tesla deaths.To our surprise the DUI deaths made up 2.5 percent of the Tesla deaths in the years 2013 - 2022. Looking at the trend of the data I would say the numbers may increase over time. Many of the deaths occur towards the later half of the data. I believe this is due to the increase in Tesla owners. Unfortunately we don't have access to that information.

# 9. Code Quality & Reproducibility

- Briefly document how someone else can reproduce your results:
- All work is in a single R Markdown file for transparency.
- We list required packages (tidyverse, dplyr, readr) in the setup chunk.
- The CSV file is read in using a relative file path.
- Knitting the Rmd fully reproduces the cleaning, plots, and summary results.

```{r session-info, echo=FALSE}
library(tidyverse)

library(dplyr)

library(readr)
Tesla_Deaths_Deaths_1_ <- read_csv("Tesla_Deaths_Deaths_1_.csv")

New_Tesla <- Tesla_Deaths_Deaths_1_ %>%
mutate(
Dscrip = if_else(
str_detect(Description, "DUI"),
"DUI",
"other"
)
)


view(New_Tesla)

New_Tesla %>%
ggplot() +
geom_freqpoly(aes(x = Year, color = Dscrip, )) +
labs(
title = "DUI",
x = "year",
y = "Description"
)



New_Tesla %>%
  count(Dscrip)


New_Tesla %>%
  count(Dscrip) %>%
  mutate(percent = n / sum(n) * 100)
```

# 10. References

List any references you used, such as:

- R Documentation (https://r4ds.hadley.nz/strings.html).
- Kaggle dataset: Tesla Accident Fatalities Analysis and Statistics (Updated 2023):
  (https://www.kaggle.com/datasets/thedevastator/tesla-accident-fatalities-analysis-and-statistic)

# Appendix (Optional)

Include extra plots, diagnostic checks, or model comparisons if needed.
